---
title: "Homework2 - Basic data types and operators"
author: "João Conde"
date: "15/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. What is the result of the following commands? Consider a possible result before executing.

```{r}
as.complex(2)
as.integer(-3.25)
as.logical("0")
as.numeric(TRUE)
as.character(11.5 + 2i)
as.numeric("ABC")
```


# 2. How do the following expressions look like in R:

# “three times ten on the power of nine”
```{r}
3*10^9
```

# “logarithm of 5”
```{r}
log(5)
```

# “integer division of 10 by 3”
```{r}
10 %/% 3
```

# “the remainder of integer division of 10 by 3”
```{r}
10 %% 3
```

# “tangent of 75 degrees”
```{r}
tan(75 * pi/180)
```


# 3. Using the if expression, check whether the result of dividing a whole number with NULL is considered to be NA, NaN or NULL.

```{r}
result <- 10/NULL

if(length(is.na(result)) != 0) {
  print("Dividing a whole number with NULL is NA")
} else if(length(is.nan(result)) != 0) {
  print("Dividing a whole number with NULL is NaN")
} else if(is.null(result)) {
  print("Dividing a whole number with NULL is NULL")
} else print("Result is none of the above (NA, NaN or NULL)")
```


# 4. Enter 5 in the variable x. Print all environment variables. Then put NULL in the x variable. Is this variable still there?

```{r}
x <- 5
ls()
x <- NULL
ls()
```

## Yes the variable is still there but with NULL value.
