---
title: "Homework6 - Packages, built-in functions and environments"
author: "João Conde"
date: "30/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Exercises

### 1. Load the following packages in the working environment: magrittr, dplyr, tidyr, ggplot2. Print the search path and check where the loaded packages are.

```{r}
packages <- c("magrittr", "dplyr", "tidyr", "ggplot2")
for(package in packages) library(package, character.only = TRUE)
search()
```


### 2. The following commands will create a vector of 20 randomly selected natural numbers from 1 to 100.

```{r}
# 20 random natural numbers from 1 to 100, with repetition

set.seed(1234)
a <- sample(1:100, 20, replace = T)

### Use the cheat sheets and/or official documentation to find built-in functions that perform the following tasks.
### Print:
### - vector a
a

### - the values of the vector a arranged in reverse order
rev(a)

### - unique values from the vector a
unique(a)

### - the values of the vector a sorted in ascending order
sort(a)
```


### 3. We mentioned that loaded packages are actually “environments”. If we want to get a direct reference to them, we need to use as.environment and the package name. Try to get a reference to the package:magrittr package in the form of an environment, and use ls to check which names are contained in it.

```{r}
ls(envir=as.environment("package:magrittr"))
```
