---
title: "Homework7 - User defined functions"
author: "João Conde"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises

### 1. R has a which function which converts a logical vector into a numeric one containing indexes where the original vector has a TRUE value (so c(T, F, F, F, F, T, F, T) becomes c(1, 6, 8)). Create a function which replicates this behaviour.

```{r}
myWhich <- function(v){
  stopifnot(is.logical(v))
  indexes <- vector()
  for(i in 1:length(v)) if(v[i]) indexes <- c(indexes, i)
  indexes
}

myWhich(c(T, F, F, F, F, T, F, T))
```


### 2. Create a factory of moment functions (moment(k)) for calculating the standardized central moment of the k-th order. Create the functions zero_moment (x) and first_moment (x) with parameter values k set to 0 and 1 accordingly. Test the functions on vector 1:1000. Compare the results given by the sd (standard deviation) function over the vector 1: 1000 and root of the first moment you have calculated.

```{r}
moment <- function(k){
  function(x){
    sum((x - mean(x))**(k+1))/length(x)
  }
}
zero_moment <- moment(0)
first_moment <- moment(1)

zero_moment(1:1000)
first_moment(1:1000)

sd(1:1000)
```


### 3. 

```{r}
Person <- function(id, surname, weight) {
    p <- list(id = id, surname = surname, weight = weight)
    class(p) <- "Person"
    p
}

print.Person <- function(p) {
    rez <- paste("ID:", p$id, ", surname:", p$surname, ", weight:", p$weight, "\n")
    cat(rez)
}

# Implement a constructor for the class Employee which inherits the Person class.
# Employee has all the attributes of the Person class as well as the superior attribute which represents a reference to a employee who is his/her superior (if such exists, otherwise it should be NA).
Employee <- function(id, surname, weight, superior){
  e <- list(id = id, surname = surname, weight = weight, superior = superior)
  class(e) <- c("Employee", "Person")
  e
}

# Create two objects of the Employee class (one superior to another) and print them with the print function. 
e1 <- Employee("ID1234", "Conde", 80, NA)
e2 <- Employee("ID4321", "Dias", 85, e1)
print(e1)
print(e2)
cat("------------------\n")

# Then implement your own version of the generic function print for the Employee class that prints employee data and his/her superior employee data (if it exists, otherwise it prints a message that there is no superior employee). Reprint both employees with the print function.
print.Employee <- function(e){
  info <- paste("ID:", e$id, ", surname:", e$surname, ", weight:", e$weight)
  cat(info)
  if(is.list(e$superior)) 
    cat(paste(" has superior ->", e$superior$id, e$superior$surname, "\n"))
  else 
    cat(" has no superior\n")
}
print(e1)
print(e2)

```

