---
title: "Homework10 - Dates and strings"
author: "João Conde"
date: "11/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises

The following tasks relate to the data set stored in the CSV file `crimeSample.csv`, which is a sample from the record of criminal incidents in the City of Philadelphia (the original data set can be found at <a href =" https://www.opendataphilly.org/ dataset / crime-incidents "> this link </a>). The original set of columns was reduced and 1000 incidents were randomly sampled from the set of all observations.

Before solving problems, load the data into the crimes data box and familiarize yourself with the data set (`str`,` head`, etc.)

```{r}
crimeDf <- read.csv("data/crimeSample.csv", stringsAsFactors = F)
str(crimeDf)
head(crimeDf)
```


1. Convert the timestamp column from character type to `POSIXct` type.
```{r}
crimeDf$Dispatch_Date_Time <- as.POSIXct(crimeDf$Dispatch_Date_Time)
str(crimeDf)
```


2. Add the following columns to the data frame: `Year`,` Month`, `Hour`. Fill in the columns with the appropriate information from the timestamp. Answer the question: in what month does the most crime occur? Which hour of the day is "most dangerous" according to the data?
```{r}
library(lubridate)
crimeDf$Year <- year(crimeDf$Dispatch_Date_Time)
crimeDf$Month <- month(crimeDf$Dispatch_Date_Time)
crimeDf$Hour <- hour(crimeDf$Dispatch_Date_Time)

# Q: In what month does the most crime occur? Which hour of the day is "most dangerous" according to the data?
## Aggregate by month
countByMonth <- table(crimeDf$Month)
monthIdx <- which.max(countByMonth)
month.abb[monthIdx]

cat("----------------------------\n")

## Aggregate by hour
countByHour <- table(crimeDf$Hour)
hour <- which.max(countByHour)
hour
```


3. Answer the question: What is the percentage of incidents where the incident description contains the word  'burglary' or 'robbery'? (tip: convert the entire crime description column to lowercase using the `tolower()` function.
```{r}
library(stringr)
crimeDf$Text_General_Code <- tolower(crimeDf$Text_General_Code)
occurrences <- nrow(crimeDf[str_detect(crimeDf$Text_General_Code, "(burglary|robbery)"),])
paste(occurrences / nrow(crimeDf) * 100, "%", sep = "")
```


4. Print any unique four-digit numbers you can find on street names where a criminal incident is recorded.
```{r}
streets <- crimeDf$Location_Block
numbers <- sapply(streets, function(street) str_extract_all(street, "\\d{4}"))
unlist(unique(numbers))
```

***
