---
title: "Exploratory analysis of 'Board Games' data set"
author: "João Conde"
date: "31st January 2020"
output:
  html_document: default
  pdf_document: default
---

## Load libraries and data set

```{r}
library(scales)
library(gridExtra)
library(tidyverse)
boardgames_tbl <- read_csv('../data/boardgames.csv')
```

## Explore data set

```{r}
head(boardgames_tbl)
glimpse(boardgames_tbl)
```

## Understanding column values' ranges and missing values

Game types:
```{r}
unique(boardgames_tbl$type)
```

Negative or null publication years:
```{r}
nrow(filter(boardgames_tbl, yearpublished <= 0))
```

Few reviewed games:
```{r}
nrow(filter(boardgames_tbl, users_rated <= 10))
```

Columns with missing values:
```{r}
colnames(boardgames_tbl)[colSums(is.na(boardgames_tbl)) > 0]
```

## Filtering dataset  

```{r}
boardgames_clean_tbl <-
  boardgames_tbl %>%
  # ignoring expansions
  filter(type == "boardgame") %>%   
  # filter negative year values and too low ones
  filter(yearpublished >= 1960 & yearpublished <= 2019) %>% 
  # eliminate games with too few reviews as the rating can be biased (e.g. one 5-star vote)
  filter(users_rated >= 10) 
```

## Top25 popular board games (based on total number of owners)

The graph below lists the top 25 games according to their popularity, based in total number of owners.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  top_n(25, total_owners) %>%
  ggplot(aes(reorder(name, total_owners), total_owners)) +
  geom_col(fill = "blue") +
  coord_flip() +
  labs(x = "Total Owners", y = "Game title") +
  ggtitle("Top25 popular games (based on total number of owners)")
```


## Maximum players average per game throughout the years

In order to evaluate if over the years game had become more multiplayer oriented, we plotted the graph below, with an average of the maximum number of players from all the games released per year and an horizontal black line as the all time average.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  group_by(yearpublished) %>%
  summarise(n = n(), total_max_players = sum(maxplayers)) %>%
  ggplot(aes(yearpublished, total_max_players/n)) +
  geom_hline(yintercept=mean(boardgames_clean_tbl$maxplayers)) +
  geom_col(fill="orange") +
  labs(x="Year", y="Average maximum players") +
  ggtitle("Average maximum players per year")
```

However, it seems board games are as multiplayer oriented as before, having very similar number of maximum players per year (excluding the outlier of 1963 which is far above the average for maximum player limit).


## Fixed or variable number of players

A simple pie chart to illustrate the percentage of games that have an exact number of players and the other games that allow for a variable number of players, within a range.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  mutate(fixed_player_size = (minplayers == maxplayers)) %>%
  summarise(nfixed = sum(fixed_player_size), nvariable = sum(!fixed_player_size)) -> slices

perc <- round((unlist(slices) / (unlist(slices)[1] + unlist(slices)[2])), 3) * 100
lbls <- c("Fixed player size", "Varying player size")
lbls <- paste(lbls, " (", perc, "%)", sep="")
pie(unlist(slices), col = c("red", "blue"), labels=lbls, main="Fixed and varying player sized games")
```


## Fixed and varying duration games

Similar to the last one, a pie chart to illustrate the percentage of games with fixed and variable durations.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  mutate(fixed_player_size = (minplaytime == maxplaytime)) %>%
  summarise(nfixed = sum(fixed_player_size), nvariable = sum(!fixed_player_size)) -> slices

perc <- round((unlist(slices) / (unlist(slices)[1] + unlist(slices)[2])), 3) * 100
lbls <- c("Fixed game duration", "Varying game duration")
lbls <- paste(lbls, " (", perc, "%)", sep="")
pie(unlist(slices), col = c("red", "blue"), labels=lbls, main="Fixed and varying duration games")
```

## minage
## total_traders
## total_wanters
## total_wishers

## Rating distribution

The graph below shows us that games' rating follow a gaussian distribution. The vertical black line represents the average of averages.

```{r, echo = FALSE}
ggplot(boardgames_clean_tbl, aes(x = average_rating)) +
  geom_vline(xintercept = mean(boardgames_clean_tbl$average_rating), color = "black") +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, fill = "red", alpha = 0.2, col = "deeppink") +
  geom_density(color="black") +
  xlab("Games Average Rating") +
  ggtitle("Ratings distribution")
```

## Ratings distribution per year since the beginning of the 21st century.

The graph below shows the distribution of game ratings per year since the beginning of the 21st century. Horizontal black line shows the average of those ratings.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  filter(yearpublished >= 2000) %>%
  mutate(year_discrete=as.factor(yearpublished)) %>%
  ggplot(aes(year_discrete, average_rating, fill=year_discrete)) +
  geom_boxplot() +
  geom_hline(yintercept=mean(boardgames_clean_tbl$average_rating), color="black")+
  theme (axis.text.x = element_text(angle = 45)) +
  labs(x = "Year", y = "Average Rating") +
  ggtitle("Ratings distribution per year since the 21st century")
```

## Popularity and ratings

In the graph below we we observe a positive correlation between the total number of owners and number of ratings and their influence on the average rating of a game. Hence, one can make the assumption that people in the community tend to prefer high-rated games. It is fair to assume that highly rated games are the most popular thus selling better.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  ggplot(aes(average_rating, users_rated)) +
  geom_point(alpha = 0.5, lwd = 0.5, color="red") +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                labels = trans_format("log10", math_format(10^.x)),
                limits = c(10,10^5)) +
  labs(x = "Average rating", y = "Number of ratings") -> g1

boardgames_clean_tbl %>%
  ggplot(aes(average_rating, total_owners)) +
  geom_point(alpha = 0.5, lwd = 0.5, color="blue") +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                labels = trans_format("log10", math_format(10^.x)),
                limits = c(10,10^5)) +
  labs(x = "Average rating", y = "Number of owners") -> g2

grid.arrange(g1, g2, top="Correlation between average rating and game popularity (number of ratings and total owners)")
```

## Game releases over time

The following bar chart and line demonstrates that since 1960 untill 2014 the amount of games released per year kept increasing. However, from 2015 onwards the amount of games released per year fell drastically.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  group_by(yearpublished) %>%
  summarise(games_published = n()) %>%
  ggplot(aes(yearpublished, games_published)) +
  geom_col() +
  geom_line(color="red") +
  labs(x = "Year", y = "Games published") +
  ggtitle("Game releases per year")
```

## Game ratings since the beginning of the 21st century

The boxplot graph below shows the average ratings of games released since the beginning of the 21st century. The horizontal black line represents the average of the ratings.

We can observe that the average of ratings has been steadily increasing.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  filter(yearpublished >= 2000) %>%
  mutate(year_discrete=as.factor(yearpublished)) %>%
  ggplot(aes(year_discrete, average_rating, color=year_discrete)) +
  geom_boxplot() +
  geom_hline(yintercept=mean(boardgames_clean_tbl$average_rating), color="black")+
  labs(x = "Year", y = "Ratings") +
  theme (axis.text.x = element_text(angle = 90)) +
  ggtitle("Ratings average per year")
```

## Complexity of games since the beginning of the 21st century

Game complexity is hard to defined and measure. However, the BoardGames data set has a 'weight' attribute, stored in the respective columns 'total_weights' and 'average_weight'. It is a score given by users that provides a reduced sense of the complexity of a game, based on their perception.

The boxplot graph below shows the average weight (or complexity) of games released since the beginning of the 21st century. The horizontal black line represents the average of the weights

We can observe that the complexity of games has been steadily increasing.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  filter(yearpublished >= 2000) %>%
  mutate(year_discrete=as.factor(yearpublished)) %>%
  ggplot(aes(year_discrete, average_weight, color=year_discrete)) +
  geom_boxplot() +
  geom_hline(yintercept=mean(boardgames_clean_tbl$average_weight), color="black")+
  labs(x = "Year", y = "Complexity (weights)") +
  theme (axis.text.x = element_text(angle = 90)) +
  ggtitle("Complexity (weights) average per year")
```


## Complexity and rating

The line below shows that higher complexity games usually have an higher rating as well. This does not hold for extremely simple and extremely complex games.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  ggplot(aes(average_weight, average_rating)) +
  geom_smooth(method = "loess") +
  labs(x = "Average complexity (weight)", y = "Average rating") +
  ggtitle("Higher complexity games tend to have higher ratings")
```

## Game comments and rating

Higher rated games appear to have more comments, hence spiking more engage from the board game community.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  ggplot(aes(total_comments, average_rating)) +
  geom_smooth() +
  labs(x = "Number of comments", y = "Average rating") +
  ggtitle("Higher rated games tend to have more community feedback (comments)")
```

Also, as expected, most of the people who comment a game, also rate it.

```{r, echo = FALSE}
boardgames_clean_tbl %>%
  ggplot(aes(total_comments, users_rated)) +
  geom_smooth() +
  labs(x = "Number of comments", y = "Number of ratings") +
  ggtitle("Users that comment almost always rate the game")
```

# Future improvements

A non explored concept was that of bayesian average. Bayesian average was later discovered during this project development's phase. Future exploratory analysis should be made using this feature.

## Average ratings and outliers (few extreme votes)

In the graph below we observe that board games with few votes tend to have overall better ratings.
A massive amount of points representing ratings are concentrated in the region of few voted games.

```{r}
ggplot(boardgames_clean_tbl, aes(users_rated, average_rating)) + 
  geom_point(alpha = 0.3) +
  xlim(0, 20000) +
  labs(x = "Number of user votes", y = "Average rating") + 
  ggtitle("Games with few votes tend to have extreme ratings")
```

Even though we filtered games with 10 or less votes, there is a *"smoother"* way of dealing with outliers. That is done using the Bayesian average. In Bayesian statistics we start out with a prior that represents our a *"priori"* assumptions. When evidence comes in we can update this prior, computing a so called posterior that reflects our updated belief. Thus if we have an unrated game we assume its average. If not, the ratings will have to convince us otherwise. As seen below, this removes outliers fairly well.

```{r}
gather(boardgames_clean_tbl, n_ratings, rating, average_rating, bayes_average_rating) %>%
  ggplot(aes(users_rated, rating)) +
  geom_point(alpha = 0.3) +
  facet_grid(. ~ n_ratings) +
  xlim(0, 20000) +
  labs(x = "Number of user votes", y = "Rating") +
  ggtitle("Removal of outliers by Bayesian average")
```

In the graph above we see that higher rated games are better distributed across different number of votes when using the bayesian average rather than the average rating.

## Recent games bias

Newer games tend to have higher ratings, due to their populatiry as seen in the graph below.

```{r}
boardgames_clean_tbl %>%
  group_by(yearpublished) %>%
  summarise(average = mean(average_rating)) %>%
  ggplot(aes(yearpublished, average)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Publishing year", y = "Average rating") +
  ggtitle("Recent games have higher ratings")
```

Using Bayesian average will also prove beneficial here since it also *smooths* the recency effect, as seen below.

```{r}
boardgames_clean_tbl %>%
  group_by(yearpublished) %>%
  summarise(average = mean(average_rating),
            bayes_average = mean(bayes_average_rating)) %>%
  gather(type, average, average, bayes_average) %>%
  ggplot(aes(yearpublished, average, bayes_average, color = type)) +
  xlim(1960, 2015) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  ggtitle('Moderation of recency effect with bayesian average')
```
