
---
title: "1 Introduction"
author: "Damir Pintar"
date: "23. august 2017"
output:
  html_document: default
---


```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'hold')
```

***

This is the first notebook that follows the coursebook "Programming in R". The notebook is designed as a **RMD (R Markdown)** file and is designed to be opened in the **RStudio** tool.

If you have opened this document in the RStudio tool on the toolbar above this text you can notice the knit button. Try to press it - you should as a result get an HTML version of the workbook that you can read with the help of a regular web browser.

The notebook follows the examples and exercises that are handled in a specific lesson. The example you see in the tutorial can be seen below:

### Example - a few simple commands of the R programming language

```{r}
3+2         # addition
log(10)    # natural logarithm!
log10(10)   # base-10 logarithm! 
sin(0.5 * pi)      # pi is one of the R constants
```

Try to execute each command separately (using the `CTRL + ENTER` key combination) or all at once (`SHIFT + ENTER`). You will see the results below the R block (if you do not see them, check if you have the option  *Chunk Output inline* turned on. Note that all the commands inside such "R Chunks" RStudio will automatically execute when converting this file to an HTML document.

Now solve the following exercise.

### Task 1.1 - Checking and editing commands

```{r}
# run the following commands by removing the comment character

getwd () # folder in which we are currently working
setwd (".") # here we can specify a new working folder if desired

```

While `=` in other programming languages is usually a default assignment operator, the operator `<-` is preferred in R. In he next exercise assign the value `5` to the variable `x`, and print this variable (with the` print` function or just by putting the variable name). Make sure you do not use the left single quotes you see in this paragraph in the program code - they are only for the purpose of generating HTML text versions of the document as part of the script. 


### Exercise 1.2 - The assignment operator

```{r}
# put `5` in the variable` x`
x <- 5
# then print the variable `x` on the screen
x
```

This is the end of a workbook, but you can freely use it to experiment with RMD files and RStudio. Some things you can try are:

- creating a new block with the `CTRL + ALT + I` key combination or choosing * Insert -> R *
- trying out different options in the setup menu (small wheel next to the Knit button)
- etc.

```{r}
print("Hello world, R")

getwd () # folder in which we are currently working
setwd ("~") # here we can specify a new working folder if desired
getwd()
```

