
---
title: "3 Vertices, matrices, lists"
author: "student"
date: "24. kolovoz 2017"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'hold')
```


***

The new vector(having more than one element) is created using the `c` (from * combine *) function.

```{r}
# numeric vector
m <- c(1, 2, 3, 4, 5)

# logic vector
v <- c(T, F, T)

# character vector
names <- c("Ivo", "Pero", "Ana")
```


***


## Exercise 3.1 - creating vectors


```{r}
# create a new vector `x` with four arbitrary elements of the following types:
# logical, real, character and integer

# print the vector content and its class screen

```


***

## Exercise 3.2  - explicit vector type conversion
```{r}
x <- c(1, T, 2L)
y <- c(1L, 2L, 3L)
z <- c(1.25, TRUE, "Ana")

# Consider a possible result and then try to make the following conversions
# vector `x` in numeric type
# vector `y` in character type
# vector `z` in a whole type
```



***


## Vector concatenation


```{r}
a <- c(1, 2, 3)
b <- c(4, 5)
c <- c(6, 7, 8) # variable can be called "c" in spite of the function c()

d <- c(a, b, c) # d is now c(1, 2, 3, 4, 5, 6, 7, 8)
```



***
In addition to the `c` function, R also offers additional convenient ways of creating new vectors:

- `:` -  "range" operator, giving the range from upper to lower bound, both included
- `seq` - sequence function, similar to the range operator but with additional options
- `rep` - replicate function, repeats the provided elements provided number of times
   
***

## Exercise 3.3  - vector creation helper functions

```{r}
# print the results of the following commands

# 1: 5
# rep(c(1, 2, 3), times = 3)
# rep(c(1, 2, 3), each = 3)
# seq(1, 5, by = 0.5)
```
  


***

Vectors can also be created by means of functions corresponding to the names of the vector types (`numeric`,` character`, etc.) whereby as a parameter we specify the desired length of the vector. This is often done as a "preparation" of the vector for subsequent filling up of real values, ie, a kind of reservation of the place in memory. What is interesting is the fact that we can also create an "empty" vector of a certain type that is still a vector, with only the length of zero (for which, for example, with the function `c` we can add elements later).

```{r}
x <- numeric(2) # vector is filled with "null" elements, in this case (0, 0)
y <- character(5)
z <- integer(0) # "empty" vector!
z <- c(z, 1) # add to the vector the element 1 (actually "merge empty vector and element 1")
```


***

Finally, to check if some vector contains a certain element we can use the operator `%in%`:

```{r, eval = F}
4 %in% seq(1, 10, 2) # returns FALSE
"d" %in% c("a", "b", "c", "d") # returns TRUE
```

***

## Operator `[`



```{r}
a <- c(2, 4, 6)
a[1] # prints a value of 2
a[2] <- 5 # element on the 2nd place becomes 5
a[5] <- 7 # at the 5th place is added 7, and the "hole" is filled with NA
a
```
 




***
 

## Exercise 3.4  - vectorization principle

```{r}
x <- seq(-5, 5, 1)
a <- 1:3
b <- 4:6


# call the abs function to calculate the absolute value
# over the vector `x` and print the result


# add vectors `a` and` b` with operator `+`
# and print the result


# multiply vectors `a` and` b` with operator `*`
# and print the result

```



***




***

## Exercise 3.5  - recycling principle 

```{r}
a <- 1:4
b <- c(1, 2)
c <- rep (5, 3)

# duplicate vector `a` and print the result

# divide vector `a` with vector` b` and print the result

# multiply vectors `a` and` c` and print the result

```

***

Now we can finally demystify the difference between "scalar" and "vector" logic operators.

- *Scala * logical operators are intended for use with single-element vectors, they return unique values of `TRUE` or `FALSE` and are suitable for use in various conditional terms.

- *Vector* logical operators use standard R's vectorization and recycling principles, ie, they are intended to work with logical vectors and as a result give a logical vector
***

## Exercise 3.6- scalar and vector logical operators

```{r}
a <- c(T, F, F)
b <- c(T, T, F)

# apply scalar and vector version of logical operator "or"
# over the `a` and` b` vectors and print the result

```



***

## Exercise 3.7- vectorization of comparison operators

```{r}
x <- 1:5
y <- seq(-10, 10, 5)

# print x and y

# print the result of the x > y command and explain the result

# print the result of the x < 3 command and explain the result

```



***

## Index vectors

We have already learned that a vector can be retrieved through a numerical index (and we did not forget the fact that the first element has an index 1). This concept can be expanded by taking multiple elements from the vector at once. which is often referred to as "slicing".

The basic principle of choosing multiple elements at once is simple -  we only need to specify the indexes of the elements we want. R offers three basic ways of crawling:

- integer- or location- based indexing
- conditional or boolean-based indexing
- label-based indexing

Which indexing we choose depends on whether we want to access the elements depending on their location, name, or condition, and each type of indexing essentially amounts to the use of a particular vector type as a parameter for the indexing operator. Such a vector is called an "index vector" because of its role.

***



### Location Indexing


*Location Indexing* is the generalization of an already familiar indexing principle where we state the ordered numbers (indices) of elements we are interested in. If we want more elements, we simply put their indices "packed" into a numeric vector.

Try solving the next task by using the appropriate numeric vectors as indexing parameters.


***

## Exercise 3.8- location-based indexing

```{r}
x <- 1:10

# print the first element of x

# print the first three vector elements x

# print the first, fifth, and seventh elements of the vector x

```

***

## Exercise 3.9 - location indexing (2)

```{r}
x <- 1:10

# answer the following questions with the help of an appropriate example

# what does index 0 return?

# what does a negative index return?

# what happens if you use an index outside of vector boundaries

```
***

Indexing is not only used to retrieve elements. By combining the indexing operator and the assignment operator we can change the vector elements (also by the principle of "multiple elements at once":




## Exercise 3.10 - location-based indexing and assignment

```{r}
a <- 1:10

# set all vector elements of `a` from the second to the eighth place to zero
# print vector `a`

b <- 1:20
b [2 * 1:5] <- 0

# Consider what the vector `b` looks like after the above command
# print the vector `b` and explain the result

```



***
### Conditional indexing

If we carefully considered the results obtained with examples with vectorized comparison operators then we can very well grasp the way *conditional indexing* works. The principle is simple - for the index vector we set a logical vector of the same length as the vector whose elements we want to retrieve. The logic vector elements determine which elements are retained (the positions where the value is `TRUE`) and which we reject (positions where the value is `FALSE`).


***

## Exercise 3.11 - conditional indexing
```{r}
x <- 1:10

# create a logical vector `y` of length 10 with an arbitrary combination of
# TRUE and FALSE values

# index the vector `x` with the` y` vector, print and explain the result


# print all vector elements `x` which are less or equal to 5
# use the appropriate expression as a logical index vector

```

***

## Exercise 3.12 - conditional indexing (2)
```{r}
y <- seq (1, 100, 7)
students <- c("Ivo", "Petra", "Marijana", "Ana", "Tomislav", "Tin")

# print all even, and then all odd vector elements of `y`

# print all vector elements from `students` which represent 3-letter names
# (note: we use the `nchar` function to count the characters in R)

```

***


### Label-based indexing

*label-based indexing* works on the principle of explicitly naming the elements we want to "keep". In order to be able to use this type of indexing we must meet the necessary prerequisite - vector elements must have predefined "names".

The vectors we used so far did not have named elements. Each element had its predefined *position* within the vector and its value but did not have any special additional identifiers. Programming language R allows you to attach names to vector elements in a very simple way - using a combination of the `names` function, the assignment operator, and the character vector with selected names. We need to make sure that the vector name is of the same length as the original vector!

***

## Exercise 3.13 - label-based indexing

```{r}
height <- c(165, 173, 185, 174, 190)
names (height) <- c("Marica", "Pero", "Josip", "Ivana", "Stipe")

# print the vector `height`


# print the height of Pero and Josip

```

***

If for some reason we want to delete the names of vector elements, simply forward `NULL` to `names`.

```{r}
names <- NULL
```

***

## Matrices and arrays 

There are several ways to create a new matrix:

- With the help of the `matrix` function to which we forward the one-dimensional vector and the desired number of rows and columns through the `nrow` and `ncol` parameters
- "manually" by setting dimensions of a one dimensional vector using the `dim` function and associating a two-element numeric vector with matrix dimensions
- "binding" rows or columns together with functions `rbind` (*row-bind*) and` cbind` (*column-bind*)



***

## Exercise 3.14 - matrix function

```{r}
x <- 1:12

# create a matrix with 3 rows and 4 columns using the `matrix` function
# print the result on the screen

# repeat the procedure but add the parameter `byrow = T` to the calling function
# print the result on the screen and compare it to the previous result

```
***

## Exercise 3.15 - dim function
```{r}
m <- 1:10

# print the result of call of the `dim` function to the vector` m`

# use the `dim` function to vector` m` with the assigment of the vector c(2, 5)

# print `m` and comment the result

# print the call results of `nrow` and` ncol` on the matrix `m`
```

***

```{r, eval = F}
m <- t(m) # transpose the matrix and store it back in the variable `m`
```


***


Finally, a matrix can be created by "gluing" rows and columns with the help of `rbind` and` cbind`. This is also a convenient way to add new rows and columns to an existing matrix.


```{r}
x <- 1:3
y <- 4:6
rbind(x,y)
```

***


## Exercise 3.16 - functions 'rbind' and 'cbind'

```{r}
a <- 1:4
b <- 5:8
c <- c(0.0)

# create a matrix `m` in which vectors `a` and `b` will be columns

# add a new row to the top of the matrix `m` with vector elements` c`
# print matrix `m`

```


***

### Matrix splicing

All the learned principles for "cutting" the vector using index vectors can be applied on matrices. The differences are as follows:

- we index each dimension individually
- first we index the lines, then the columns, and divide the index vectors by comma
- If we want "all rows" or "all columns" we simply omit this index vector( but we still use a comma)

```{r, eval = F}
# assume that `m 'is a matrix of 3 x 5, with column names from `a` to `e`

m[1, 2:5] # first line, all columns from second to fifth
m[c(F, T, T), c("a", "b")] # second and third rows, columns `a` and` b`
m[,] # all rows and all columns (same as just `m`)

```




## Exercise 3.17 - matrix splicing

```{r}
m <- matrix (1:30, 6, 5, T)
colnames (m) <- c("a", "b", "c", "d", "e")

# print all elements of the matrix m from the second to the fourth line
# and from the third to the fifth column

# set all elements in column "c" to zero
# and then print the first two lines of matrix `m`

# print only column "d"

# rewrite column "d", but add the `drop = FALSE` parameter when indexing
# Separate the parameter with a comma (as if it was a "third" indexing dimension)


```

***

Finally, although we will not deal in detail with the arrays, we will show the example of the program code that creates a three-dimensional array and then uses standard cutting principles we encountered with the vectors and matrices.

```{r, eval = F}
myArray <- array(1:24, dim = c(2, 3, 4)) # array of dimension 2 x 3 x 4

myArray[, 1:2, 3, drop = FALSE] # print all rows, first and second columns
                                        # 3rd "layer," with array type retention
```

***

## Lists

The list is an element in R used as a "universal container" of any data. Unlike the vector (ie, the concept of vector as we initially defined it), the list may contain different types of data or, more often, sets of different types of data.

We create the list with the `list` function by which we add a string of names of elements and their content. These elements can be anything, even other lists.

```{r, eval = F}
mojaLista <- list(a = 1, b = 2:100, c = list(x = 1, y = 2))
```


***

## Exercise 3.18 - list creation

```{r}
# create a new list called `stuff` that will have the following elements
# element called `numbers' with integers from 1 to 3
# element called `letters' with letters 'A' and 'B'
# nameless element with logical vector `c(T, F)`
# element called `titles' with the concent of 'Mr', 'Mrs' and 'Ms'


# print the `stuff` variable

```


***

## Exercise 3.19 - list structure

```{r}
# print the structure of the `stuff` list

```

***

## List = vector?

At the beginning of this lesson, we have said that the principle "all is a vector" is very important in R and that the vectors actually have arranged sets of elements of the same type. From this one could conclude that this fact is not valid for the lists - they obviously contain elements of different types. But the real answer is - the lists *are* actually vectors, and the definition is actually still consistent. That is, all the elements of the list are actually a small single-element lists, so all elements are formally of the same type.

***

## Exercise 3.20 - list element type
```{r}
# print the first element of the list `stuff`

# check its type

```

***

## Operator `[[`

So, we've proved that the list elements are actually a small list, as seen from printing the item itself, as well as checking its type. It may seem that the elements above created lists should be vectors, since we have created the list by "stacking" different vectors, however in the process of list creation all these elements are wrapped in single-element lists.

Often we do not want to work with a list element as a "small list", but want to have it in its "original" form. For this we use the operator `[[`, ie the operator of "double angular brackets".

***

## Exercise 3.21 - operator `[[`
```{r}
# print the first element of the list `stuff` using the operator `[[`


# check its type

```

***

## Operator `$`


The aforementioned operator is most often used to retrieve the selected element of the list defined by the number or (if defined) by the element name. In this approach, we must use the syntax such as `list[[name_element]]` symbol, which is somewhat clumsy for typing. Because of this, R offers an alternative way of accessing the list elements by their name using the `$` operator, ie `list$name_element`.


***

## Exercise 3.22  - operator `$`

```{r}
# print the `letters` element of the `stuff` list
# using `[[` the operator 


# print the `letters` element of the `stuff` list
# using the `$` operator 

```

***

The lists are an extremely popular type of object in R, as they represent a universal template for more complex data structures, including more complex objects in the narrower sense (as we will see later). The list is also the "foundation" for the most popular and most commonly used element of the R-language: the data frame - which we will learn in the next lesson.

Finally, we learn how to add an element to the list. This is easiest to do using the aforementioned operator `$` - such as `list$newElement <- newElementContents`. The element is deleted by assigning the value `NULL`.


***

## Exercise 3.23  - adding list elements

```{r}
# in the `stuff` list add the `evenNumbers` element which contains
# all even numbers from 1 to 100

# delete the third element from the list


# print the `stuff` list
```


